
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(demography)
library(ggplot2)
library(timetk)
```

State names list
```{r}

states <- c("AL","AR","AZ","CA","CO","CT","DE","FL","GA","IA", "ID",
            "IL","IN","KS","KY","LA","MA", "MD","ME","MI", "MN",
            "MO","MS","MT","NC", "ND", "NE","NH","NJ","NM","NV","NY",
            "OH","OK","OR","PA","RI","SC","SD","TN", "TX","UT",
            "VA","VT","WA","WI","WV","WY")

states_help <- states
#Alaska es Hawaii-t kivettem mert nincsenek szarazfoldi szomszedos allamaik, tovabbi meg DC is benne volt az adatok kozott igy azt is kivettem mert az nem allam
```

Global variables
```{r}

min_year_global <- 1969 #exp szamok 1969-tol indulnak
max_year_global <- 2019

train_test_border <- 2005

n <- length(states)
```


Create a list from the mx data
```{r}

states_mx <- paste(states,"_bltper_1x1.txt", sep="")

list_of_path_mx <- file.path("~/Desktop/TDK/bltper_1x1", states_mx)

```

mx
```{r}
mx <- list()

for (i in 1:n){
  name_of_states <- paste("mxt_", states[i], sep='')
  
  helper1 = read.table(list_of_path_mx[i], skip = 2, header=T) 
  
   helper1 = helper1 %>%
    filter(Age != "110+") %>%
    filter(Year >= min_year_global & Year <= max_year_global)
  
  helper1$Age = as.numeric(helper1$Age)
  helper1$Year = as.numeric(helper1$Year)
  
  helper1 = helper1 %>%
    filter(Age <= 89)
  
  mx[[i]] <- helper1[, c("Year", "Age", "mx")]
}

names(mx) = paste("mxt_", states, sep='')
```

exp
source:
https://seer.cancer.gov/popdata/download.html#single
https://seer.cancer.gov/popdata/popdic.html
```{r}
exp_data <- "~/Desktop/TDK/us.1969_2023.singleages.through89.90plus.adjusted.txt"

Exp <- read.table(exp_data, stringsAsFactors = FALSE, colClasses = "character", fill = TRUE)

Year_chr         <- substr(Exp$V1, 1, 4)
State            <- substr(Exp$V1, 5, 6)
State_FIPS_chr  <- substr(Exp$V1, 7, 8)
County_FIPS_chr <- substr(Exp$V1, 9, 11)
Race_chr        <- substr(Exp$V2, 1, 1)
Origin_chr      <- substr(Exp$V2, 2, 2)
Sex_chr          <- substr(Exp$V2, 3, 3)
Age_chr          <- substr(Exp$V2, 4, 5)
Pop_chr          <- substr(Exp$V2, 6, 14)

Year         <- as.integer(Year_chr)
State_FIPS  <- as.integer(State_FIPS_chr)
County_FIPS <- as.integer(County_FIPS_chr)
Race        <- as.integer(Race_chr)
Origin      <- as.integer(Origin_chr)
Sex          <- as.integer(Sex_chr)
Age          <- as.integer(Age_chr)
Pop          <- as.integer(Pop_chr)

Exp<- data.frame(
  State = State,
  State_FIPS = State_FIPS,
  Year = Year,
  County_FIPS = County_FIPS,
  Race = Race,
  Origin = Origin,
  Sex = Sex,
  Age = Age,
  Pop = Pop,
  stringsAsFactors = FALSE
)

Exp <- Exp[Exp$Year <= 2019 & Exp$Age <= 89 & Exp$State !="AK" & Exp$State !="HI" & Exp$State !="DC" & Exp$State !="KR",]

# Aggregate: sum population across sexes for each Year x State x Age
Exp <- aggregate(Pop ~  State + Year + Age, data = Exp, FUN = function(x) sum(x, na.rm = TRUE))
unique(Exp$State)
# sort result
Exp <- Exp[order(Exp$State, Exp$Year, Exp$Age), ]
#sub state sorting
exposures <- list()

for(s in states){
  sub <- Exp[Exp$State == s, c("Year","Age","Pop")]
  exposures[[s]] <- sub
}

names(exposures) = paste("exp_", states, sep='')
```


```{r}
mortality_data_raw <- bind_rows(mx, .id = "state")
mortality_data_raw <- mortality_data_raw %>%
  mutate(state = sub("^mxt_", "", state))                  

mortality_data_raw <- mortality_data_raw %>% 
  rename(states = state)

#num_steps_fc <- max_year_global-train_test_border+1

helper <- mortality_data_raw
helper$exp <- Exp$Pop
helper$dx <- helper$mx*helper$exp+0.5
helper$mx <- helper$dx/helper$exp

mortality_data <- helper[,c(1,2,3,4)]

mx_new <- list()

for (i in 1:n){
  name_of_states <- paste("mxt_", states[i], sep='')
  
  mortality_data_help = mortality_data[mortality_data$states == states_help[i],]
  
  mx_new[[i]] <- mortality_data_help[, c("Year", "Age", "mx")]
}

names(mx_new) = paste("mxt_", states_help, sep='')

```


mx Train and test set
```{r}
mx_train_LC <- list()
mx_test_LC <- list()

for (i in 1:n){
 
  mx_train_LC[[i]] <- mx_new[[i]] %>% filter(Year < train_test_border)
  mx_test_LC[[i]] <- mx_new[[i]] %>% filter(Year >= train_test_border)
}

names(mx_train_LC) = paste("mxt_train_LC_", states, sep='')
names(mx_test_LC) = paste("mxt_test_LC_", states, sep='')

```

exp Train and test set
```{r}

exp_train_LC <- list()
exp_test_LC <- list()

for (i in 1:n){
  
  exp_train_LC[[i]] <- exposures[[i]] %>% filter(Year < train_test_border)
  exp_test_LC[[i]] <- exposures[[i]] %>% filter(Year >= train_test_border)
}

names(exp_train_LC) = paste("Ext_train_LC_", states, sep='')
names(exp_test_LC) = paste("Ext_test_LC_", states, sep='')
```

LC:
```{r,warning=FALSE}

#Data
mxt <- list()
Ext <- list()

#Models
lc_svd <- list()

#LC estimated parameters
kappaSVD_dt_t <- list()
alpha_x_dt <- list()
beta_x_dt <- list()

#Parameters
min_year <- 1969
max_year <- 2019

min_age <- 0
max_age <- 89

nr <- (max_age-min_age+1)
nc <- (train_test_border-min_year)

for(i in 1:n){
  
 mxt[[i]] <- matrix(mx_train_LC[[i]]$mx,nrow = nr, ncol = nc)
 Ext[[i]] <- matrix(exp_train_LC[[i]]$Pop, nrow = nr, ncol = nc)

 
 #Build LC models
 lc <- demogdata(mxt[[i]], Ext[[i]], min_age:max_age, min_year:(train_test_border-1), "mortality", "-", "Total")
 lc_svd[[i]] <- lca(lc, adjust = "dt")
 
 #Save parameters into vectors
 kappaSVD_dt_t[[i]] <- as.vector(lc_svd[[i]]$kt)
 alpha_x_dt[[i]] <- as.vector(lc_svd[[i]]$ax)
 beta_x_dt[[i]] <- as.vector(lc_svd[[i]]$bx)
}

names(lc_svd) = paste("LC_SVD_", states[1:n], sep="")

#plot(lc_svd[[10]])
```


Kappa vectors into long format
```{r}

kappa_df_base_LC_SVD_dt <- data.frame(matrix(ncol = n, nrow = length(kappaSVD_dt_t[[1]])))

for (i in 1:n){
  
  kappa_df_base_LC_SVD_dt[,i] <- kappaSVD_dt_t[[i]]
}

colnames(kappa_df_base_LC_SVD_dt) <- states

years <- seq(from = min_year_global, to = train_test_border-1, by = 1)
kappa_df_LC_SVD_dt <- cbind(years, kappa_df_base_LC_SVD_dt)


kappa_df_long_LC_SVD_dt <- data.frame(kt = unlist(kappa_df_base_LC_SVD_dt, use.names = FALSE))
countr <- rep(states, length(years))

kappa_df_long_LC_SVD_dt <- cbind(year = rep(years, n), states = sort(countr), kappa_df_long_LC_SVD_dt)
```

Check kappa t constraints. -- Ha van kiigazítás akkor sum kappa = 0 feltételnek nem kell teljesülnie! Ez csak a kiigazítatlan verziónál fontos.
```{r}

kappa_df_long_LC_SVD_dt %>%
  group_by(states) %>%
  summarize(sum_kt = sum(kt)) %>%
  print(n = 50)
```

bx - Collect them into a data frame - Rows: years, Columns: Countires
```{r}

beta_df_base_LC <- data.frame(matrix(ncol = n, nrow = length(beta_x_dt[[1]])))

for (i in 1:n){
  
  beta_df_base_LC[,i] <- beta_x_dt[[i]]
}

colnames(beta_df_base_LC) <- states

ages <- seq(from = min_age, to = max_age, by = 1)
beta_df_LC <- cbind(ages, beta_df_base_LC)
```

Transform the data frame into long format - This is ideal to panel models
```{r}

beta_df_long_LC <- data.frame(bx = unlist(beta_df_base_LC, use.names = FALSE))
countr <- rep(states, length(ages))

beta_df_long_LC_SVD_dt <- cbind(states = sort(countr), age = rep(ages, n), beta_df_long_LC)
```

Check beta x constraints. 
```{r}

beta_df_long_LC_SVD_dt %>%
  group_by(states) %>%
  summarize(sum_bx = sum(bx)) %>%
  print(n = 50)
```

ax - Collect them into a data frame - Rows: years, Columns: Countires
```{r}

alpha_df_base_LC <- data.frame(matrix(ncol = n, nrow = length(alpha_x_dt[[1]])))

for (i in 1:n){
  
  alpha_df_base_LC[,i] <- alpha_x_dt[[i]]
}

colnames(alpha_df_base_LC) <- states

ages <- seq(from = min_age, to = max_age, by = 1)
alpha_df_LC <- cbind(ages, alpha_df_base_LC)
```

Transform the data frame into long format - This is ideal to panel models
```{r}

alpha_df_long_LC <- data.frame(ax = unlist(alpha_df_base_LC, use.names = FALSE))
countr <- rep(states, length(ages))

alpha_df_long_LC_SVD_dt <- cbind(states = sort(countr), age = rep(ages, n), alpha_df_long_LC)
```

Save the parameters
```{r}

write.csv(kappa_df_long_LC_SVD_dt,
          "~/Desktop/TDK/output/LeeCarter_SVD_Kappat_dt_Total.csv",
          row.names=FALSE)

write.csv(beta_df_long_LC_SVD_dt,
          "~/Desktop/TDK/output/LeeCarter_bx_SVD_dt_Total.csv",
          row.names=FALSE)

write.csv(alpha_df_long_LC_SVD_dt,
          "~/Desktop/TDK/output/LeeCarter_ax_SVD_dt_Total.csv",
          row.names=FALSE)
```

